---
title: "R Notebook"
output: html_notebook
---

```{r}
library(data.table)
library(DataExplorer)
library(skimr)
```


#Summary data for ct_schools overall school "Outcome Rate" report card for all public schools in CT 

Mean outcome rate has been declining in all categories but Charter. Regional School averages are above Charters, Magnet and broad average.

```{r}

## 202 School Districts had mean outcome rate decline from 77.8 to 73.8 over 4 years
school_summary[school_name=="District"][
            ][,.(mean(outcome_rate_pct,na.rm=TRUE),
                 .N),
                  by=year]

# Average outcome rate and achievement gap all districts during 2017 in decreasing order
school_summary[school_name=="District",
           .( outcome_rate_pct=mean(outcome_rate_pct,na.rm=TRUE),
              achievement_gap_flag=sum(achievement_gap_flag,na.rm = TRUE)),
                by=district_name][
                    ][order(-outcome_rate_pct)]

#There are 24 public charters with average outcome rate of 64 which is below the broader average. The mean outcome rate for charter has risen over 4 years
school_summary[school_org_type== "Public Charter Schools"][
           ][,.(.N,
                mean(outcome_rate_pct,na.rm=TRUE),
                mean(achievement_gap_flag,na.rm=TRUE)),
                          by=year]

#72 schools have "Region" in name or are in a "Regional School District" 
# with an average outcome rate 77 which has been declining 
school_summary[(school_org_type== "Regional Schools" |
              str_detect(district_name,"Region")) & 
             school_name!="District"][
                ][,.(.N,mean(outcome_rate_pct,na.rm=TRUE),
                      mean(achievement_gap_flag,na.rm=TRUE)),
                          by=year]

#71 schools have "Magnet" in name with an average outcome rate of 66.8 slightly lower over 4 years
school_summary[str_detect(school_name,"Magnet")][
  ][,.(.N,
       mean(outcome_rate_pct,na.rm=TRUE),
       mean(achievement_gap_flag,na.rm=TRUE)),
                  by=year]

```


#Table of summary mean test scores by year

Total measured population is close to 900 schools. Mean test scores have increased slightly over the period from 2014-2017.

```{r}

#Add grp based on school type
metrics_prep[,grp:=.GRP,by=c("is_elem","is_middle","is_high")]
metrics_prep[,.(is_elem,is_middle,is_high,grp)]

#Summary of grp and year based on test_score
map=setdiff(names(metrics_prep),drops)
cube(metrics_prep[grp %in% c(3,4,2,5)],
     j=c(list(count=.N),lapply(.SD,mean,na.rm=TRUE)),
     .SDcols="test_score",
     by=c("grp","year"))

```

# Summary stats for charter vs non-charter by group

```{r}
cols=c("grp","year","pct_eic","cost_student","pct_female","free","egl_pop","tanf_pop","empl","pct_min","se_yes","st","test_score","pct_teach","total","dist_enroll","is_charter","is_magnet","is_regional")

#all schools year and type for year 2017
cube(metrics_prep[ year=="2017",.(grp,year,test_score,is_charter)],j=c(list(count=.N),mn=mean(test_score,na.rm=TRUE)),by=c("grp","is_charter","year"))[,dcast(.SD,grp+year~ is_charter,fun.aggregate=mean,value.var=c("mn","count"))]

#Charter test_scores by year and type
groupingsets(metrics_prep[grp %in% c(3,4,2,5),.(grp,year,test_score,is_charter,is_regional,is_magnet)],j=c(list(count=.N),mn=mean(test_score,na.rm=TRUE)),by=c("grp","year","is_charter"),sets=list("is_charter","grp"))[,dcast(.SD,grp~is_charter,value.var="mn")]

```

# Skimr summary

```{r}

skim_to_list(metrics_prep[,.SD,.SDcols=patterns(paste(cols,collapse="|"))])
```


# Data Explorer

```{r}

create_report(metrics_prep[,.SD,.SDcols=patterns(paste(cols,collapse="|"))],
              y="test_score",
              config = configure_report(
              add_plot_prcomp = FALSE)
  )


```




