---
title: "R Notebook"
output: html_notebook
---

```{r "setup", message=FALSE, include=FALSE, results='hide'}
require("knitr")
require("MarkdownReports")
opts_knit$set(root.dir = "~/Desktop/David/Projects/ct_schools")
```

# Libraries

```{r "libraries", echo=FALSE, message=FALSE, warning=FALSE}
library(data.table)
library(stringr)
```

# Load CT Municipal Financial Indicators since 2001

```{r echo=FALSE, message=FALSE, warning=FALSE}
ct_mfi <-
  readRDS("/Users/davidlucey/Desktop/David/Projects/ct_muni/ct_mfi.RDS")

#Build smaller data.table from ct_mfi of education attributes
mfi_edu <- 
  ct_mfi[fisc_year_end > 2001][
    ][, .(
          year = fisc_year_end,
          municipality,
          education,
          enrollmt,
          egl,
          population = x_population,
          tanf_recipients,
          empl
          )][
            ][, `:=`(
              town = str_to_title(as.character(municipality)),
              year = as.character(year),
              egl_pop = egl / (population - enrollmt),
              tanf_pop = tanf_recipients / population
            )][
              ][, municipality := NULL]

setcolorder(mfi_edu, c(1, 8, 2:7, 9:10))

#Inspect
mfi_edu[, lapply(.SD, uniqueN)][, melt(.SD)][][order(value)]
mfi_edu[, lapply(.SD, function(x)
  sum(is.na(x)) / .N)][][, melt(.SD)][order(-value)]

rm(ct_mfi)

```

# Load IRS data and build percent earned income credit (pct_eic) variable

```{r echo=FALSE, message=FALSE, warning=FALSE}
irs <- readRDS("/Users/davidlucey/Desktop/David/Projects/irs/irs.RDS")[state=="CT"]
zips <- fread("~/Desktop/David/Projects/CT_data/ct_zips.csv")

pct_eic <- 
  irs[year %in% as.character(c(2014:2017)) & 
      !zipcode %in% c("00000","99999") & state=="CT"][
        ][,.(zip=zipcode,
             year,
             agi_level,
             n59660,
             n1,a59660)][
          ][,(pc_eic=sum(n59660,na.rm=TRUE)/sum(n1,na.rm=TRUE)),
              by=.(year,zip)][
            ][zips[,zip:=zipcode::clean.zipcodes(zip)],
                    on="zip"][
              ][,.(year,town=str_to_title(municipality),V1)][
                ][,.(pct_eic=sum(V1,na.rm=TRUE)),
                     by=.(year,town)]

pct_eic[, lapply(.SD, uniqueN)][,melt(.SD)][
  ][order(value)]
pct_eic[, lapply(.SD, function(x) sum(is.na(x))/.N)][
  ][,melt(.SD)][order(-value)]

rm(irs)
```


# Load all 1094 schools and engineer school type binary variables
# Remove address, phone and convert "city" variable to "town"

```{r "school_directory", echo=FALSE, message=FALSE, warning=FALSE}
#List of schools in CT 
school_dir <- fread("Education_Directory.csv")
school_dir <- janitor::clean_names(school_dir)

#school_dir <- 
school_dir <- 
  school_dir[,zipcode:=str_extract(zipcode,"^\\d{5}")][
             ][zips[,.(city=str_to_title(municipality),
                       zipcode=zipcode::clean.zipcodes(zip))],
                                  on="zipcode"][
               ][!is.na(school_name)]

##Add features for kind of school by school & remove location attributes other than zip
school_dir <- 
  school_dir[organization_type != "Public School Districts"][
        ][,interdistrict_magnet :=
            as.integer(fifelse(!is.na(interdistrict_magnet),0,1))][
          ][, `:=`(is_magnet = 
                     as.integer(fifelse(str_detect(organization_type, "Magnet") | str_detect(school_name,"Magnet"),1L,0L)),
                    is_regional = 
                     as.integer(fifelse(str_detect(organization_type, "Region|Unified|Coop|Learn") | str_detect(school_name,"Region|Unified|Coop|Learn"),1L,0L)),
                    is_charter = 
                     as.integer(fifelse(organization_type=="Public Charter Schools",1,0)),
                    is_correct=
                     as.integer(str_detect(school_name,"Correct")),
                    organization_code =NULL,
                    address= NULL,
                    phone=NULL,
                    location=NULL,
                    town=city)][
                    ][,`:=`(is_pk=as.integer(fifelse(pre_kindergarten>0,1,0)),
                            is_elem=as.integer(fifelse(grade_3>0,1,0)),
                            is_middle=as.integer(fifelse(grade_8>0,1,0)),
                            is_high=as.integer(fifelse(grade_10>0,1,0)),
                            city=NULL
                            )][
                        ][,.SD,.SDcols=!patterns("grade_|kindergarten|organization|open")][
                          ][, district_size := .N, by=district_name]

school_type <- fread("specializedschoolreports.csv",
                     skip = 3,
                     select=c(1:2))
names(school_type)[1:2] <- c("district_name","school_name")
school_type$is_magnet_1 <-1L

school_dir <- 
  school_type[school_dir, 
              on=c("district_name","school_name")][
              ][,is_magnet_1 :=fifelse(!is.na(is_magnet_1),1,0)][
                ][,is_magnet:=
                    as.integer(fifelse((is_magnet==1 |interdistrict_magnet==1 |is_magnet_1==1),1,0))][
                    ][,`:=`(is_magnet_1=NULL, interdistrict_magnet=NULL)]

#Summary of unique levels and NA's
school_dir[, lapply(.SD, uniqueN)][,melt(.SD)][order(value)] 
school_dir[, lapply(.SD, function(x) (sum(is.na(x)))/.N)][,melt(.SD)][order(-value) & value>0][,.(variable,format(value,scientific=FALSE))]

```

# Load schools summary data for all public schools and districts from 2014-2017
# This data includes the full 12 metrics which are measured for each school system

```{r "school_summary", echo=FALSE, message=FALSE, warning=FALSE}
#Download all school summary data 2004-2017 into ct_schools data.table
subdir <- "./school_edsight/"
school_summary <- rbindlist(lapply(list.files(subdir,full.names = TRUE),fread))
school_summary <- janitor::clean_names(school_summary)

school_summary[,`:=`(year=as.character(fall_of_year),
                     district_name=rptng_district_name)][
                ][,`:=`(fall_of_year=NULL,
                        rptng_district_name=NULL)]

#drop code columns and add town of school
school_summary <- school_summary[,.SD,.SDcols=!patterns("code")][
  ][,town:= trimws(str_remove(str_remove(district_name,"School"),"District"))][
    ][,year:=as.character(year)]

#function to replace char class "."  with NA
clean_dot <- function(x) {
  x <- fifelse(x %in% c("", "."),NA_character_,x)
  x
}

# use clean_dot formula with specified columns and convert those columns to NA
cols <- names(school_summary)[which(str_detect(names(school_summary),"^ind|pct$|^final|^grad|flag$"))]
school_summary <- 
  school_summary[,(cols):=lapply(.SD,clean_dot), .SDcols=cols][
      ][,(cols):=lapply(.SD,as.numeric),.SDcols=cols]

# summary data about number of unique values and number of NA's in each data
school_summary[, lapply(.SD, uniqueN)][,melt(.SD)][order(-value)] 
school_summary[, lapply(.SD, function(x) (sum(is.na(x)))/.N)][,melt(.SD)][order(-value)]

```


# Load school detailed performance data by group

```{r "school_perf", echo=FALSE, message=FALSE, warning=FALSE}
# Download detailed test data by school
subdir <- "./school_performance/"
school_perf <- rbindlist(lapply(list.files(subdir,
                                               full.names = TRUE),
                                    fread,
                                    skip=7,
                                    select=c(2,4:12)),
                             use.names = TRUE,
                             fill=TRUE,
                             idcol="year")

#Convert idcol to date
school_perf[,year:=as.character(year+2013)]

#school_perf <- school_perf 
school_perf <- janitor::clean_names(school_perf)

#Clean asterisk function 
clean_asterisk <- function(x) {
  x <- fifelse(x == "*",NA_character_,x)
  x
}

#Convert "*" to NA on specified columns
cols <- c(6:11)
school_perf <- school_perf[,(cols):=lapply(.SD,clean_asterisk), .SDcols=cols]

# summary data about number of unique values and number of NA's in each data
school_perf[, lapply(.SD, uniqueN)][
  ][,melt(.SD)][
    ][order(value)] 
school_perf[, lapply(.SD, function(x) (sum(is.na(x)))/.N)][
  ][,melt(.SD)][order(-value) & value>0][
    ][,.(variable,format(value,scientific=FALSE))]

school_perf
```


# Upload ct_enrollment data for all schools 2007-18
# CT has 10% fewer students than 2007

```{r "school_enroll", echo=FALSE, message=FALSE, warning=FALSE}
#Load granulaar school enrollment data for 2007-2018 
subdir <- "./school_enroll/"
files <- gtools::mixedsort(list.files(subdir))
school_enroll <- rbindlist(lapply(paste0(subdir,files), fread, na.strings=''), idcol = "year")[,-c(3,5)]

#Convert idcol to date
school_enroll[,year:=as.character(year+2006)]

#Change names to match other datasets
setnames(school_enroll,
         old=c("year","District","School","Count"),
         new=c("year","district_name","school_name","enroll"))

#Roll down district name in spreadsheet and change "*" char to NA
school_enroll[,2] <- lapply(school_enroll[,2], zoo::na.locf)

#Set enroll var equal to * to NA
school_enroll$enroll <- fifelse(school_enroll$enroll=="*",NA_character_,school_enroll$enroll)

#Convert ct_enroll to numeric
school_enroll$enroll <- as.numeric(school_enroll$enroll)

#Total ct public school enrollment
school_enroll[,sum(enroll,na.rm=TRUE),by=year]

#Years filtered 2014-17
school_enroll <- school_enroll[year %in% as.character(2014:2017)]

#Summary of unique levels and NA's
school_enroll[, lapply(.SD, uniqueN)][,melt(.SD)][order(value)] 
school_enroll[, lapply(.SD, function(x) (sum(is.na(x)))/.N)][,melt(.SD)][order(-value) & value>0][,.(variable,format(value,scientific=FALSE))]


```


# Upload and merge school demographic data over 4 categories (race, Special Ed, free lunch, ESL) from 015-2018 into single ct_schools_dems file
```{r "school_demos", echo=FALSE, message=FALSE, warning=FALSE}

##Function to take .csv datasets by year, rbind and clean "*" char to NA
upload_school_data <- function(dataset) {
  years <- c("2014","2015","2016","2017","2018")
  string<- paste0("school_",dataset,"/",dataset,years,".csv")
  x <- lapply(string,fread,na.strings="")
  names(x) <- years
  x <- rbindlist(x,idcol = TRUE)[,-c(3,5)]
  #x[,2][x[,2] == ""] <- NA
  x[,2] <- lapply(x[,2], zoo::na.locf)
  x[,4:ncol(x)][x[,4:ncol(x)] == "*"] <- NA
  return(x)
}

#Upload separate testing data by race, Special Ed, free lunch and ESL and merge into school_dems
dataset <- c("race","SE","lunch","esl","gender")
x <- lapply(dataset,upload_school_data)
names(x) <- dataset
names(x[[2]])[4:5] <- paste0("SE_",names(x[[2]])[4:5])
names(x[[4]])[4:5] <- paste0("esl_",names(x[[4]])[4:5])

#combine datasets from list of each type of data into ct_school_dems
school_dems <- merge(x[[1]],x[[2]],by=c(".id","School","District"))
school1_dems <- merge(x[[3]],x[[4]],by=c(".id","School","District"))
school_dems <- merge(school_dems,school1_dems,by=c(".id","School","District"))
school_dems <- merge(school_dems,x[[5]],by=c(".id","School","District"))
school_dems <- school_dems[,.SD,.SDcols=!patterns(".x|.y")]
school_dems[,4:ncol(school_dems)][school_dems[,4:ncol(school_dems)] == "0"] <- 0
#school_dems[which(is.na(school_dems[,4:ncol(school_dems)])),4:ncol(school_dems)] <- "0"

#Equalize names for joining
school_dems <- janitor::clean_names(school_dems)
names(school_dems)[1:3] <- c("year","school_name","district_name")

#Convert Variables to numeric
cols <- c(4:19)
school_dems[, (cols):=lapply(.SD,as.numeric),.SDcols=cols]

#Aggregate free and reduced price lunch
school_dems <- 
  school_dems[,lunch:=fcoalesce(free+reduced,total-non_subsidized)][
    ][,`:=`(sub=NULL,sub1=NULL,lunch=fifelse(lunch>0,lunch,0))]

#Summary of unique levels and NA's
school_dems[year %in% as.character(2014:2017), lapply(.SD, uniqueN)][,melt(.SD)][order(value)] 
school_dems[year %in% as.character(2014:2017), lapply(.SD, function(x) (sum(is.na(x)))/.N)][,melt(.SD)][order(-value) & value>0][,.(variable,format(value,scientific=FALSE))]

#Raname data and clean global env
rm(school1_dems)

```



##Teachers by School
```{r "teachers", echo=FALSE, message=FALSE, warning=FALSE}
teachers <- fread("FTEStaffing.csv",skip=3,na=".")
teachers <- janitor::clean_names(teachers)
teachers[,`:=`(school_name=school,
                district_name=district)][
          ][,`:=`(school=NULL,
                  district=NULL)]

teachers[,c(20:21)][teachers[,c(20:21)] ==""] <- NA
teachers[,c(20:21)] <- lapply(teachers[,c(20:21)], zoo::na.locf)

#teachers <- teachers[unique(schools[,.(school_name,town)]), on="school_name"]

names(teachers)[3:18] <- str_extract(names(teachers)[3:18],"[[:digit:]]{4}")

teach <- 
  teachers[str_detect(assignment_category,"Teachers") & 
           !str_detect(school_name, "-- District Level --"),-19][
            ][,melt(.SD,
                    id.vars=c("district_name","school_name"),
                    measure.vars= c("2014","2015","2016","2017"),
                    variable.name="year",
                    value.name="teachers")][
              ][,year:=as.character(year)][
                ][,.(tot_teach=sum(teachers,na.rm=TRUE)),
                    by=.(year,school_name,district_name)]

staff <-  
  teachers[!str_detect(school_name, "-- District Level --")][
            ][,melt(.SD,
                    id.vars=c("district_name","school_name"),
                     measure.vars=c("2014","2015","2016","2017"),
                            variable.name="year",
                            value.name="staff")][
                        ][,year:=as.character(year)][
                          ][,.(tot_staff=sum(staff,na.rm=TRUE)),
                                  by=.(year,school_name,district_name)]

teach <- teach[staff, on=c("year","school_name","district_name")]

teach[, lapply(.SD, uniqueN)][,melt(.SD)][order(value)] 
teach[, lapply(.SD, function(x) (sum(is.na(x)))/.N)][,melt(.SD)][order(-value) & value>0][,.(variable,format(value,scientific=FALSE))]

rm(staff,teachers)
```

                

#Edsight state funding is missing regional districts and Hartford-- NEEDS WORK

```{r "school_expenses", echo=FALSE, message=FALSE, warning=FALSE}
school_fin <- fread("FinanceComparison.csv",select=c(1:3,5,7))[-c(1:2),]
names(school_fin) <- c("district_name","expense_type",c(2014:2016))

#Roll down open space below district name in spreadsheet
school_fin[,1][school_fin[,1] == ""] <- NA
#ct_school_fin[,2:13][ct_school_fin[,2:13] == "N/A"] <- NA
school_fin[,1] <- lapply(school_fin[,1], zoo::na.locf)

#Convert columns to numeric
cols=c(3:5)
school_fin[,(cols) :=(lapply(.SD,as.numeric)),.SDcols=cols]

#Impute missing 2017 as +3.5% 2016 spending by district
school_fin[,`2017`:=round(`2016`*1.035,0),by=district_name]

#Extract Total Expenditure and Teacher pay
school_fin <- school_fin[expense_type %in% c("Total Expenditures","Instructional Staff and Services") ][
              #&!str_detect(expense_type,"Family Academy District")][
                ][,melt(.SD,
                        id.vars=c("expense_type","district_name"),
                        variable.name="year",
                        value.name="school_cost")][
                  ][year %in% c(2014:2017)]

#Convert year var to chr to match other datasets
school_fin$year <- as.character(school_fin$year)

#Compare MFI to Edsight Total Expenditures report
school_fin[expense_type=="Total Expenditures"][
  ][,.(colSums(.SD,na.rm=TRUE),.N),.SDcols=c(4),year]
mfi_edu[as.integer(year) %in% c(2014:2017)][
                      ][,.(mfi=sum(education,na.rm=TRUE),.N),by=year]

#Town percent funding by state
school_fin[district_name=="Hartford School District" & 
           expense_type == "Total Expenditures" & 
           !str_detect(expense_type,"Family Academy District")][
            ][,melt(.SD,id.vars=c("expense_type","district_name"))][,.(state= sum(value,na.rm=TRUE)),by=variable], mfi_edu[year %in% c(2005:2015) & town=="Hartford",.(mfi=sum(education)),by=year][order(year,decreasing=FALSE)])[,.(percent=mfi/state),by=year]

#Missing districts including Hartford
school_fin[ expense_type == "Total Expenditures" & 
              !str_detect(expense_type,"Family Academy District"),
              melt(.SD, id.vars=c("expense_type","district_name"))][is.na(value) & !str_detect(district_name,"Charter")]

school_fin[,.(sum(school_cost,na.rm=TRUE),.N),by=year]
```


# Cost per student from ctschoolfinance.org

```{r "school_cost", echo=FALSE, message=FALSE, warning=FALSE}
school_cost <- readxl::read_excel("2016-17-2017-18-NCEP-Comparison.xlsx")
school_cost <- janitor::clean_names(school_cost)
school_cost <- as.data.table(school_cost[1:266,c(2:4)])

school_cost1 <- readxl::read_excel("2014-15-2015-16-NCEP-Comparison.xlsx")
school_cost1 <- janitor::clean_names(school_cost1)
school_cost1 <- as.data.table(school_cost1)[1:166,c(2:4)]

school_cost <- school_cost1[school_cost,on="district_name"][
  ][,district_name:=str_to_title(district_name)]

names(school_cost)[2:5] <- str_extract(names(school_cost)[2:5],"\\d{4}")

cols <- names(school_cost)[str_detect(names(school_cost),"\\d{4}")]
school_cost <- school_cost[, (cols):= lapply(.SD,as.numeric),.SDcols=cols]

school_cost <-
  school_cost[,melt(.SD,
            id.vars="district_name",
            measure.vars=c(2:5),
            variable.name="year")][!is.na(district_name)& !str_detect(district_name,"Statewide")][
        ][,`:=`(cost_student=round(as.numeric(value),digits=0),
                district_name=ifelse(str_detect(district_name,"District"),
                                     str_c("Regional School District ",
                                           str_extract(district_name,"\\d{1,2}")),
                                     district_name),
                town=ifelse(!str_detect(district_name,"District"),
                            district_name,
                            NA))][
             ][,-c("value")]

school_cost
rm(school_cost1)
```


