---
title: "R Notebook"
output: html_notebook
---

```{r "setup", message=FALSE, include=FALSE, results='hide'}
require("knitr")
require("MarkdownReports")
opts_knit$set(root.dir = "~/Desktop/David/Projects/ct_schools")
```

Libraries

```{r "libraries", echo=FALSE, message=FALSE, warning=FALSE}
library(data.table)
library(stringr)
```

Load CT Municipal Financial Indicators since 2001

```{r echo=FALSE, message=FALSE, warning=FALSE}
ct_mfi <-
  readRDS("/Users/davidlucey/Desktop/David/Projects/ct_muni/ct_mfi.RDS")

#Build smaller data.table from ct_mfi of education attributes
mfi_edu <-
  ct_mfi[fisc_year_end > 2001][, .( # filter rows
    year = fisc_year_end, # select vars
    municipality,
    education,
    enrollmt,
    egl,
    population = x_population,
    tanf_recipients,
    empl
  )][, `:=`( #mutate vars
    town = str_to_title(as.character(municipality)), 
    year = as.character(year),
    egl_pop = egl / (population - enrollmt),
    tanf_pop = tanf_recipients / population
  )][, municipality := NULL] # drop var

 # Change col order
setcolorder(mfi_edu, c(1, 8, 2:7, 9:10))

# Inspect unique and NA's
mfi_edu[, lapply(.SD, uniqueN)][, melt(.SD)][order(value)]
mfi_edu[, lapply(.SD, function(x)
  sum(is.na(x)) / .N)][, melt(.SD)][order(-value)]

# Clean up genv
rm(ct_mfi)

```

Load IRS data and build percent earned income credit (pct_eic) variable

```{r echo=FALSE, message=FALSE, warning=FALSE}
irs <- readRDS("/Users/davidlucey/Desktop/David/Projects/irs/irs.RDS")[state=="CT"]
zips <- fread("~/Desktop/David/Projects/CT_data/ct_zips.csv")

pct_eic <-
  irs[year %in% as.character(c(2014:2017)) &
      !zipcode %in% c("00000", "99999") & # filter rows
      state == "CT"][, .( # select variables
        zip = zipcode,
        year, 
        agi_level,
        n59660,
        n1,
        a59660)][, .(
        pc_eic = sum(n59660, na.rm = TRUE) / sum(n1, na.rm = TRUE)), 
          by = .(year, zip)][ #summarize pct_eic by year, zip
        zips[, zip := zipcode::clean.zipcodes(zip)], on = "zip"][ 
          #join zips to find town of zip
          ][, .(
        year, # select and mutate new varis
        town = str_to_title(municipality), 
        V1)][, .(
        pct_eic = sum(V1, na.rm = TRUE)), 
        by = .(year, town)] #summarize pct_eic by year/town

# Inspect unique and NA's
pct_eic[, lapply(.SD, uniqueN)][,melt(.SD)][
  ][order(value)]
pct_eic[, lapply(.SD, function(x) sum(is.na(x))/.N)][
  ][,melt(.SD)][order(-value)]

# Clean up
rm(irs)
```


Load all 1094 schools and engineer school type binary variables. 
Remove address, phone and convert "city" variable to "town"

```{r "school_directory", echo=FALSE, message=FALSE, warning=FALSE}
#List of schools in CT 
school_dir <- fread("Education_Directory.csv")
school_dir <- janitor::clean_names(school_dir)

school_dir <-
  school_dir[, zipcode := str_extract(zipcode, "^\\d{5}")][
    #extract 5-digit zip from school_dir as zipcode
  ][zips[, .(
    city = str_to_title(municipality),
    zipcode = zipcode::clean.zipcodes(zip)
    )], on = "zipcode"][
    #join zipcode packaage to convert zip to city
      ][!is.na(school_name)] # drop na school_names from join

# Add features for kind of school by school & 
# remove location attributes other than zip
school_dir <- 
  school_dir[ # filter rows
    organization_type != "Public School Districts"][
  ][, `:=`( # mutate variables
    interdistrict_magnet = 
      as.integer(fifelse(!is.na(interdistrict_magnet),0,1)),
    is_magnet = 
      as.integer(fifelse(str_detect(organization_type, "Magnet") |
                           str_detect(school_name,"Magnet"),1L,0L)),
    is_regional = 
      as.integer(fifelse(str_detect(organization_type, 
                                    "Region|Unified|Coop|Learn") |
      str_detect(school_name,"Region|Unified|Coop|Learn"),1L,0L)),
    is_charter = 
      as.integer(fifelse(organization_type=="Public Charter Schools",1,0)),
    is_correct=as.integer(str_detect(school_name,"Correct")),
    organization_code =NULL,
    address= NULL,
    phone=NULL,
    location=NULL,
    town=city,
    is_pk=as.integer(fifelse(pre_kindergarten>0,1,0)),
    is_elem=as.integer(fifelse(grade_3>0,1,0)),
    is_middle=as.integer(fifelse(grade_8>0,1,0)),
    is_high=as.integer(fifelse(grade_10>0,1,0)),
    city=NULL)][
  ][,.SD, # select cols based on regex pattern
    .SDcols=!patterns("grade_|kindergarten|organization|open")][
  ][, district_size := .N, # mutate district_size
    by=district_name]

# Import school type
school_type <- 
  fread("specializedschoolreports.csv",
                     skip = 3,
                     select=c(1:2))
names(school_type)[1:2] <- 
  c("district_name","school_name")
school_type$is_magnet_1 <-1L

# Join school_type and school_dir
school_dir <-
  school_type[school_dir,
    on = c("district_name", "school_name")][
      # intermediate mutate is_magnet_1
  ][, is_magnet_1 := fifelse(!is.na(is_magnet_1), 1, 0)][
      # mutate variables
  ][, is_magnet := as.integer(fifelse((
          is_magnet == 1 | 
          interdistrict_magnet == 1 | 
          is_magnet_1 == 1), 1, 0))][
      # drop intermediate variables
  ][, `:=`(is_magnet_1 = NULL, 
           interdistrict_magnet = NULL)]

# Inspect unique and NA'ss
school_dir[, lapply(.SD, uniqueN)][
  ][,melt(.SD)][order(value)] 
school_dir[, lapply(.SD, function(x) (sum(is.na(x)))/.N)][
  ][,melt(.SD)][order(-value) & value>0][
    ][,.(variable,format(value,scientific=FALSE))]

```

Load schools summary data for all public schools and districts from 2014-2017. This data includes the full 12 metrics which are measured for each school system

```{r "school_summary", echo=FALSE, message=FALSE, warning=FALSE}
#Download all school summary data 2004-2017 into ct_schools data.table
subdir <- "./school_edsight/"
school_summary <- 
  rbindlist(lapply(list.files(subdir,full.names = TRUE),fread))
school_summary <- 
  janitor::clean_names(school_summary)

# Mutate variables
school_summary[, `:=`( # mutate variables
    year = as.character(fall_of_year),
    district_name = rptng_district_name)][
  ][, `:=`( # drop variables
    fall_of_year = NULL,
    rptng_district_name = NULL
    )]

# drop code columns and add town of school
school_summary <-
  school_summary[, .SD, 
    # select cols with regex pattern "code"
    .SDcols = !patterns("code")][
  ][, `:=`( # mutate vars
    town = 
      trimws(str_remove(str_remove(district_name, "School"), "District")),
    year = as.character(year)
    )]

#function to replace char class "."  with NA
clean_dot <- function(x) {
  x <-
    fifelse(x %in% c("", "."), NA_character_, x)
  return(x)
}

# use clean_dot formula with specified columns and convert those columns to NA
cols <-  # cols to select
  names(school_summary)[
    which(str_detect(names(school_summary),
      "^ind|pct$|^final|^grad|flag$"))]

# Clean "*" cells and convert to numeric
school_summary <-
  # Run clean_dot on selected cols
  school_summary[, (cols) := lapply(.SD, clean_dot),
    .SDcols = cols][
  # Convert all selected cols to numeric
  ][, (cols) := lapply(.SD, as.numeric), 
    .SDcols = cols]

# Inspect unique and NA's
school_summary[, lapply(.SD, uniqueN)][
  ][, melt(.SD)][order(-value)]
school_summary[, lapply(.SD, function(x)
  (sum(is.na(x))) / .N)][
  ][, melt(.SD)][order(-value)]

```


Load school detailed performance data by group

```{r "school_perf", echo=FALSE, message=FALSE, warning=FALSE}
# Download detailed test data by school
subdir <- "./school_performance/"
school_perf <- 
  rbindlist(lapply(list.files(
    subdir,
    full.names = TRUE),
    fread,
    skip=7,
    select=c(2,4:12)),
    use.names = TRUE,
    fill=TRUE,
    idcol="year")

# Convert idcol to actual year
school_perf[, year := as.character(year + 2013)]

# Clean names
school_perf <- janitor::clean_names(school_perf)

# Clean asterisk function
clean_asterisk <- function(x) {
  x <- fifelse(x == "*", NA_character_, x)
  return(x)
}

#Convert "*" to NA on specified columns
cols <- c(6:11)
school_perf <-
  school_perf[, (cols) := lapply(.SD, clean_asterisk), 
    .SDcols = cols]

# Inspect unique and NA's
school_perf[, lapply(.SD, uniqueN)][
  ][,melt(.SD)][
    ][order(value)] 
school_perf[, lapply(.SD, function(x) (sum(is.na(x)))/.N)][
  ][,melt(.SD)][order(-value) & value>0][
    ][,.(variable,format(value,scientific=FALSE))]

```


Upload ct_enrollment data for all schools 2007-18 CT has 10% fewer students than 2007

```{r "school_enroll", echo=FALSE, message=FALSE, warning=FALSE}
# Load granular school enrollment data for 2007-2018 
subdir <- "./school_enroll/"
files <- gtools::mixedsort(list.files(subdir))
school_enroll <-
  rbindlist(lapply(paste0(
    subdir,files),
    fread, 
    na.strings=''),
    idcol = "year")[,-c(3,5)]

# Convert idcol to actual year
school_enroll[, year := as.character(year + 2006)]

# Change names to match other datasets
setnames(
  school_enroll,
  old = c("year", "District", "School", "Count"),
  new = c("year", "district_name", "school_name", "enroll")
)

#Roll down district name in spreadsheet and change "*" char to NA
school_enroll[, 2] <- 
  lapply(school_enroll[, 2], zoo::na.locf)

#Set enroll var equal to * to NA
school_enroll$enroll <- 
  fifelse(school_enroll$enroll=="*",
          NA_character_,
          school_enroll$enroll)

# Convert ct_enroll to numeric
school_enroll$enroll <-
  as.numeric(school_enroll$enroll)

# Total ct public school enrollment
school_enroll[, sum(enroll, na.rm = TRUE), by = year]

# Years filtered 2014-17
school_enroll <-
  school_enroll[year %in% as.character(2014:2017)]

# Inspect unique and NA's
school_enroll[, lapply(.SD, uniqueN)][
  ][, melt(.SD)][order(value)]
school_enroll[, lapply(.SD, function(x) (sum(is.na(x))) / .N)][
  ][, melt(.SD)][order(-value) & value > 0][
    ][, .(variable, format(value, scientific = FALSE))]


```


Upload and merge school demographic data over 4 categories (race, Special Ed, free lunch, ESL) from 015-2018 into single ct_schools_dems file

```{r "school_demos", echo=FALSE, message=FALSE, warning=FALSE}

# Function to take .csv datasets by year, rbind and clean "*" char to NA
upload_school_data <- function(dataset) {
  years <- c("2014", "2015", "2016", "2017", "2018")
  string <- paste0("school_", dataset, "/", dataset, years, ".csv")
  x <- lapply(string, fread, na.strings = "")
  names(x) <- years
  x <- rbindlist(x, idcol = TRUE)[, -c(3, 5)]
  x[, 2] <- lapply(x[, 2], zoo::na.locf)
  x[, 4:ncol(x)][x[, 4:ncol(x)] == "*"] <- NA
  return(x)
}

#Upload separate testing data by race, Special Ed, free lunch and ESL and merge into school_dems
dataset <- c("race", "SE", "lunch", "esl", "gender")
school_raw <- lapply(dataset, upload_school_data)
names(school_raw) <- dataset # set names
names(school_raw[[2]])[4:5] <- 
  paste0("SE_", names(school_raw[[2]])[4:5])
names(school_raw[[4]])[4:5] <- 
  paste0("esl_", names(school_raw[[4]])[4:5])

#combine datasets from list of each type of data into ct_school_dems
school_dems <- 
  merge(school_raw[[1]], school_raw[[2]], 
    by = c(".id", "School", "District"))
school1_dems <- 
  merge(school_raw[[3]], school_raw[[4]], 
    by = c(".id", "School", "District"))
school_dems <-
  merge(school_dems, school1_dems, 
    by = c(".id", "School", "District"))
school_dems <-
  merge(school_dems, school_raw[[5]], 
    by = c(".id", "School", "District"))
school_dems <- 
  school_dems[, .SD, .SDcols = !patterns(".x|.y")]
school_dems[, 4:ncol(school_dems)][
  ][school_dems[, 4:ncol(school_dems)] == "0"] <-
  0

# Equalize names for joining
school_dems <-
  janitor::clean_names(school_dems)
names(school_dems)[1:3] <-
  c("year", "school_name", "district_name")

# Convert select variables to numeric
cols <- c(4:19)
school_dems[, (cols) := lapply(.SD, as.numeric), .SDcols = cols]

# Aggregate free and reduced price lunch
school_dems <-
  school_dems[, 
    lunch := fcoalesce(free + reduced, total - non_subsidized)][
  ][, `:=`(
      sub = NULL,
      sub1 = NULL,
      lunch = fifelse(lunch > 0, lunch, 0)
      )]

# Inspect unique and NA's
school_dems[year %in% as.character(2014:2017),
            lapply(.SD, uniqueN)][
  ][, melt(.SD)][order(value)]
school_dems[year %in% as.character(2014:2017),
    lapply(.SD, function(school_raw) (sum(is.na(school_raw))) / .N)][
  ][, melt(.SD)][
    ][order(-value) & value > 0][
      ][, .(variable, format(value, scientific = FALSE))]

#Clean global env
rm(school1_dems)
```



##Teachers by School
```{r "teachers", echo=FALSE, message=FALSE, warning=FALSE}
teachers <- fread("FTEStaffing.csv", skip = 3, na = ".")
teachers <- janitor::clean_names(teachers)
teachers[, `:=`(
  school_name = school,
  district_name = district)][
  ][, `:=`(
    school = NULL,
    district = NULL
    )]

# Replace "" with NA and roll down NA's
teachers[, c(20:21)][teachers[, c(20:21)] == ""] <- NA
teachers[, c(20:21)] <-
  lapply(teachers[, c(20:21)], zoo::na.locf)

names(teachers)[3:18] <-
  str_extract(names(teachers)[3:18], "[[:digit:]]{4}")

teach <-
  teachers[
    str_detect(assignment_category, "Teachers") &
    !str_detect(school_name, "-- District Level --"),-19][
    ][, melt(
      .SD,
      id.vars = c("district_name", "school_name"),
      measure.vars = c("2014", "2015", "2016", "2017"),
      variable.name = "year",
      value.name = "teachers"
      )][
  ][, year := as.character(year)][
  ][, .(tot_teach = sum(teachers, na.rm = TRUE)),
    by = .(year, school_name, district_name)]

staff <-
  teachers[!str_detect(school_name, "-- District Level --")][
  ][, melt(
      .SD,
      id.vars = c("district_name", "school_name"),
      measure.vars = c("2014", "2015", "2016", "2017"),
      variable.name = "year",
      value.name = "staff"
    )][
 ][, year := as.character(year)][
 ][, .(tot_staff = sum(staff, na.rm = TRUE)),
   by = .(year, school_name, district_name)]

teach <- 
  teach[staff, on = c("year", "school_name", "district_name")]

teach[, lapply(.SD, uniqueN)][
  ][, melt(.SD)][order(value)]
teach[, lapply(.SD, function(x) (sum(is.na(x))) / .N)][
  ][, melt(.SD)][order(-value) & value > 0][
    ][, .(variable, format(value, scientific = FALSE))]

rm(staff,teachers)
```

                

# Edsight state funding is missing regional districts 

```{r "school_expenses", echo=FALSE, message=FALSE, warning=FALSE}
school_fin <- fread("FinanceComparison.csv",select=c(1:3,5,7))[-c(1:2),]
names(school_fin) <- c("district_name","expense_type",c(2014:2016))

#Roll down open space below district name in spreadsheet
school_fin[,1][school_fin[,1] == ""] <- NA
school_fin[,1] <- lapply(school_fin[,1], zoo::na.locf)

#Convert columns to numeric
cols=c(3:5)
school_fin[,(cols) :=(lapply(.SD,as.numeric)),.SDcols=cols]

#Impute missing 2017 as +3.5% 2016 spending by district
school_fin[,`2017`:=round(`2016`*1.035,0),by=district_name]

#Extract Total Expenditure and Teacher pay
school_fin <- 
  school_fin[
    expense_type %in% 
      c("Total Expenditures",
        "Instructional Staff and Services")][
  ][,melt(
    .SD,
    id.vars=c("expense_type","district_name"),
    variable.name="year",
    value.name="school_cost")][
  ][year %in% c(2014:2017)]

#Convert year var to chr to match other datasets
school_fin$year <- as.character(school_fin$year)

#Compare MFI to Edsight Total Expenditures report
school_fin[expense_type=="Total Expenditures"][
  ][,.(colSums(.SD,na.rm=TRUE),.N),.SDcols=c(4),year]
mfi_edu[as.integer(year) %in% c(2014:2017)][
                      ][,.(mfi=sum(education,na.rm=TRUE),.N),by=year]

#Town percent funding by state
#school_fin[district_name=="Hartford School District" & 
#           expense_type == "Total Expenditures" & 
#           !str_detect(expense_type,"Family Academy District")][
#            ][,melt(.SD,id.vars=c("expense_type","district_name"))][,.(state= sum(value,na.rm=TRUE)),by=variable], mfi_edu[year %in% c(2005:2015) & town=="Hartford",.(mfi=sum(education)),by=year][order(year,decreasing=FALSE)])[,.(percent=mfi/state),by=year]

#Missing districts including Hartford
school_fin[
  expense_type == "Total Expenditures" &
    !str_detect(expense_type, "Family Academy District")][
  ][, melt(.SD, 
           id.vars = c("expense_type", "district_name"))][
  ][is.na(value) & 
    !str_detect(district_name, "Charter")]

school_fin[,.(sum(school_cost,na.rm=TRUE),.N),by=year]
```


# Cost per student from ctschoolfinance.org

```{r "school_cost", echo=FALSE, message=FALSE, warning=FALSE}

school_cost <- # import excel
  readxl::read_excel("2016-17-2017-18-NCEP-Comparison.xlsx")
school_cost <- # clean names
  janitor::clean_names(school_cost)
school_cost <- # select rows/cols and convert to data.table
  as.data.table(school_cost[1:266, c(2:4)])

school_cost1 <- # import excel
  readxl::read_excel("2014-15-2015-16-NCEP-Comparison.xlsx")
school_cost1 <- # clean names
  janitor::clean_names(school_cost1)
school_cost1 <- # select rows/cols and convert to data.table
  as.data.table(school_cost1)[1:166, c(2:4)]

school_cost <- 
  school_cost1[school_cost, # join school_cost & school_cost1
               on = "district_name"][
  ][, # mutate district_name to title
    district_name := str_to_title(district_name)]

# seelect names to year
names(school_cost)[2:5] <-
  str_extract(names(school_cost)[2:5], "\\d{4}")

# Convert selected cols to numeric
cols <- 
  names(school_cost)[str_detect(names(school_cost),"\\d{4}")]
school_cost <- 
  school_cost[, (cols):= lapply(.SD,as.numeric),.SDcols=cols]

# 
school_cost <-
  school_cost[,
    melt(.SD,
      id.vars="district_name",
      measure.vars=c(2:5),
      variable.name="year")][
    !is.na(district_name) & 
    !str_detect(district_name,"Statewide")][
  ][,`:=`( # mutate vars
    cost_student=
      round(as.numeric(value),digits=0),
    district_name=
      ifelse(str_detect(district_name,"District"),
             str_c("Regional School District ",
                   str_extract(district_name,"\\d{1,2}")),
             district_name),
    town=ifelse(!str_detect(district_name,"District"),
                NA)
    )][
  ][,-c("value")]

# Clean up
rm(school_cost1)
```


