---
title: "R Notebook"
output: html_notebook
---


```{r}
library(data.table)
library(WVPlots)
library(ggplot2)
library(patchwork)
```

# Cols needed for graphs

```{r}
cols <- c("lunch","test_score","year","pct_min","se_yes","type")
```

# Hex plots of key explanatory vars vs test_score

```{r}


p1 <- HexBinPlot(metrics_prep[year=="2017",.SD,.SDcols=patterns(paste(cols,collapse="|"))], "lunch", "test_score", "Test Scores as a function of Free Lunch, Percent Minority & Special Ed") +
  geom_smooth(color="black", se=FALSE, na.rm = TRUE)
p2 <- HexBinPlot(metrics_prep[year=="2017" & pct_min <1], "pct_min", "test_score", "") 
  geom_smooth(color="black", se=FALSE, na.rm = TRUE)
p3 <- HexBinPlot(metrics_prep[year=="2017" & !is.na(test_score) & !is.na(se_yes)], "se_yes", "test_score", "") +
  geom_smooth(color="black", se=FALSE, na.rm = TRUE)

p1  + p2 + p3 + plot_layout(ncol = 1, widths=c(1,1,1), heights = c(3,3,3))


```


# Scatter Histograms of lunch vs test_score for Elementary, Middle and High schools

```{r}

unlist(lapply(list("Elem","Middle","High"),function(x) WVPlots::ScatterHist(
  metrics_prep[type==x,.SD,.SDcols=patterns(paste(cols,collapse="|"))], "lunch", "test_score",
  "Expected Test Score as function Free Lunch",
  smoothmethod = "lm",
  point_alpha = 0.025,
  estimate_sig = TRUE)))
```

# Two variable scatter Histograms for free lunch vs pct_minority versus test_score by school type

```{r}
plot_data<- metrics_prep[year=="2017" & pct_min<1,.(lunch,pct_min,test_score,type)]
plot_data <- plot_data[complete.cases(plot_data)]

unlist(lapply(list("Elem","Middle","High"), function(x) WVPlots::ScatterHistN(plot_data[type==x], "lunch", "pct_min", "test_score", nclus=4, title="Combined Effect of Free Lunch and Minority Students on Test Score")))

#ggplotly(p)

```

# Two variable scatter Histograms for cost_student vs student teacher ratio (st) versus test_score by school type

```{r}
plot_data <- metrics_prep[ cost_student<100000,.(cost_student,st,test_score,type)] 
plot_data <- plot_data[complete.cases(plot_data)]

unlist(lapply(list("Elem","Middle","High"), function(x) WVPlots::ScatterHistN(plot_data[type==x], "st", "cost_student", "test_score", nclus=4, title="Example Joint Distribution")))

```


```{r}
plot_data <- metrics_prep[,.(is_charter,lunch,pct_min,test_score,type,grp=.GRP),by=c("is_elem","is_middle","is_high")]
plot_data <- plot_data[complete.cases(is_charter)]

unlist(lapply(list("Elem","Middle","High"),WVPlots::ScatterHistC(plot_data[x==type], xvar="lunch", yvar= "test_score", cvar="is_charter", title="Free Lunch vs Pct Minority with Charters shown in Red")))

```


# Shadow histogram of test_score by school type for 2017

```{r}

plot_data <- metrics_prep[year=="2017",.(is_charter,lunch,pct_min,test_score,type,grp=.GRP),by=c("is_elem","is_middle","is_high")]

ShadowHist(plot_data[grp %in% c(2:4)], "test_score","grp",
                         title = "Highest Scores in School Type")


```



```{r}
WVPlots::ShadowPlot(metrics_prep,  "type", "is_charter",title = "Shadow plot of eye colors by sex")
```


