---
title: "R Notebook"
output: html_notebook

---


```{r message=FALSE, warning=FALSE, include=FALSE}

library(data.table)
library(patchwork)
library(ggplot2)
library(knitr)
library(DT)

```

# Available Test Counts and Indices for Stamford Public Schools by Student Group

```{r 'data_table', message=FALSE, warning=FALSE, include=FALSE}

dt <- datatable(school_perf[stringr::str_detect(district_name,"Stamford")][,.(year,school=school_name,student_grp=student_group,ela_count,ela_perf_idx=ela_performance_index,math_count,math_perf_idx=math_performance_index)],extensions = c("Buttons" , "FixedColumns"),
           filter = 'top')
```


```{r echo=FALSE, message=FALSE, warning=FALSE}
dt
```


# Stamford Elementary School Math Aggregated Test Scores Compared by Gender

```{r, echo=FALSE, message=FALSE, warning=FALSE}

school_perf[str_detect(district_name, "Stamford") &
                        !str_detect(school_name,"High") &
                        category== "Gender" & 
                        !str_detect(school_name,"Middle"),
         .(mean_ela=mean(as.numeric(ela_performance_index),na.rm=TRUE),
                  mean_math=mean(as.numeric(math_performance_index),na.rm=TRUE)),
         by=.(school_name,year,student_group)][
  ,ggplot(.SD,aes(as.integer(year),mean_math,color=school_name)) +
    geom_line() + 
    ggtitle("Stamford Primary Math")+ 
    ylab("Score") +  
    xlab("Year") +
    scale_x_continuous(breaks =  c(2014,2015,2016,2017,2018)) +
    facet_wrap(~student_group)+
    theme(legend.position = 'topright')+
    theme_bw()]

```


# Stamford Elementary School English Aggregated Test Scores Compared by Gender

```{r, echo=FALSE, message=FALSE, warning=FALSE}
school_perf[str_detect(district_name, "Stamford") &
                       !str_detect(school_name,"High") & 
                       category== "Gender" & 
                       !str_detect(school_name,"Middle"),
                     .(mean_ela=mean(as.numeric(ela_performance_index),na.rm=TRUE),
                       mean_math=mean(as.numeric(math_performance_index),na.rm=TRUE)),
                     by=.(school_name,year,student_group)][
  ,ggplot(.SD,aes(as.integer(year),mean_ela,color=school_name)) +
    geom_line() +
    ggtitle("Stamford Primary English")+ 
    ylab("Score") +
    xlab("Year") + 
    scale_x_continuous(breaks =  c(2014,2015,2016,2017,2018)) +
    facet_wrap(~student_group) +
    theme(legend.position = "topright")+
    theme_bw()]

```

# Stamford, Norwalk and Connecticut Averages (excluding Stamford and Norwalk) for Math by Grade

```{r, echo=FALSE, message=FALSE, warning=FALSE}

#Performance declines by grade level
p <- school_perf[category=="Grade" & district_name=="Stamford School District"][
  ][,.(math=mean(as.numeric(math_performance_index),na.rm=TRUE)),
        by=.(student_group,year)][
          ][,ggplot(.SD,aes(as.integer(year),
                            math,
                            color=student_group))+
              geom_line()+ 
              ggtitle("Stamford Math")+ 
              ylab("Score") + 
              xlab("Year")+ 
              scale_x_continuous(breaks =  c(2014,2015,2016,2017,2018))+
              scale_y_continuous(limits=c(45,70))+
              theme_bw()+
              theme(legend.position = 'none')]  

p1 <- school_perf[category=="Grade" & district_name=="Norwalk School District"][
  ][,.(math=mean(as.numeric(math_performance_index),na.rm=TRUE)),
        by=.(student_group,year)][
          ][,ggplot(.SD,aes(as.integer(year),
                            math,
                            color=student_group))+
              geom_line()+ 
              ggtitle("Norwalk Math")+ 
              ylab("Score") + 
              xlab("Year")+ 
              scale_x_continuous(breaks =  c(2014,2015,2016,2017,2018))+
              scale_y_continuous(limits=c(45,70))+
              theme_bw()+
              theme(legend.position = 'bottom')]  


p2 <- school_perf[category=="Grade" & !district_name %in% c("Stamford School District", "Norwalk School District")][
        ][,.(math=mean(as.numeric(math_performance_index),na.rm=TRUE)),
                by=.(student_group,year)][
          ][,ggplot(.SD,aes(as.integer(year),
                            math,
                            color=student_group)) + 
              geom_line()+ 
              ggtitle("Connecticut Math")+ 
              ylab("Score") + 
              xlab("Year")+ 
              scale_x_continuous(breaks =  c(2014,2015,2016,2017,2018))+
              scale_y_continuous(limits=c(45,70))+
              theme_bw()+
              theme(legend.position = 'none')] 

p + p1 + p2 + plot_layout(ncol=3)  
```

# Stamford, Norwalk and Connecticut Averages (excluding Stamford and Norwalk) for English by Grade

```{r, echo=FALSE, message=FALSE, warning=FALSE}

#Performance declines by grade level
p <- school_perf[category=="Grade" & district_name=="Stamford School District"][
  ][,.(english=mean(as.numeric(ela_performance_index),na.rm=TRUE)),
        by=.(student_group,year)][
          ][,ggplot(.SD,aes(as.integer(year),
                            english,
                            color=student_group))+
              geom_line()+ 
              ggtitle("Stamford English")+ 
              ylab("Score") + 
              xlab("Year")+ 
              scale_x_continuous(breaks =  c(2014,2015,2016,2017,2018))+
              scale_y_continuous(limits=c(45,70))+
              theme_bw()+
              theme(legend.position = 'none')] 

p1 <- school_perf[category=="Grade" & district_name=="Norwalk School District"][
        ][,.(english=mean(as.numeric(ela_performance_index),na.rm=TRUE)),
                by=.(student_group,year)][
          ][,ggplot(.SD,aes(as.integer(year),
                            english,
                            color=student_group)) + 
              geom_line()+ 
              ggtitle("Norwalk English")+ 
              ylab("Score") + 
              xlab("Year")+ 
              scale_x_continuous(breaks =  c(2014,2015,2016,2017,2018))+
              scale_y_continuous(limits=c(45,70))+
              theme_bw()+
              theme(legend.position = 'bottom')] 

p2 <- school_perf[category=="Grade"& !district_name %in% c("Stamford School District", "Norwalk School District")][
        ][,.(english=mean(as.numeric(ela_performance_index),na.rm=TRUE)),
                by=.(student_group,year)][
          ][,ggplot(.SD,aes(as.integer(year),
                            english,
                            color=student_group)) + 
              geom_line()+ 
              ggtitle("Connecticut English")+ 
              ylab("Score") + 
              xlab("Year")+ 
              scale_x_continuous(breaks =  c(2014,2015,2016,2017,2018))+
              scale_y_continuous(limits=c(45,70))+
              theme_bw()+
              theme(legend.position = 'none')] 

p + p1 + p2 + plot_layout(ncol=3)  
```


# Trajectory of Math Scores for Grades 3, 4 and 5 in Stamford Elementary Schools

```{r, echo=FALSE, message=FALSE, warning=FALSE}
p <- school_perf[category=="Grade" & 
                       student_group == "03" &
                       str_detect(district_name,"Stamford")][
  ][,.(math=mean(as.numeric(math_performance_index),na.rm=TRUE)),
          by=.(school_name,year)][
          ][,ggplot(.SD,aes(as.integer(year),
                            math,
                            color=school_name))+
              geom_line()+ 
              ggtitle("3rd Grade Math")+
              ylab("Score") + 
              xlab("Year")+ 
              scale_x_continuous(breaks = c(2014,2015,2016,2017,2018))+
              scale_y_continuous(limits=c(60,80))+
              theme_bw() +
              theme(legend.position = 'none')]

p1 <- school_perf[category=="Grade" & 
                       student_group == "04" &
                       str_detect(district_name,"Stamford")][
  ][,.(math=mean(as.numeric(math_performance_index),na.rm=TRUE)),
          by=.(school_name,year)][
          ][,ggplot(.SD,aes(as.integer(year),
                            math,
                            color=school_name))+
              geom_line()+ 
              ggtitle("4th Grade Math")+
              ylab("Score") + 
              xlab("Year")+ 
              scale_x_continuous(breaks = c(2014,2015,2016,2017,2018))+
              scale_y_continuous(limits=c(60,80))+
              theme_bw()+
              theme(legend.position = 'bottom')] 

p2 <- school_perf[category=="Grade" & 
                       student_group == "05" &
                       str_detect(district_name,"Stamford")][
  ][,.(math=mean(as.numeric(math_performance_index),na.rm=TRUE)),
          by=.(school_name,year)][
          ][,ggplot(.SD,aes(as.integer(year),
                            math,
                            color=school_name))+
              geom_line()+ 
              ggtitle("5th Grade Math")+
              ylab("Score") + 
              xlab("Year")+ 
              scale_x_continuous(breaks = c(2014,2015,2016,2017,2018))+
              scale_y_continuous(limits=c(60,80))+
              theme_bw() +
              theme(legend.position = 'none')] 

p + p1 + p2 + plot_layout(ncol=3)
```

# Trajectory of English Scores for Grades 3, 4 and 5 in Stamford Elementary Schools Only

```{r, echo=FALSE, message=FALSE, warning=FALSE}
p <- school_perf[category=="Grade" & 
                       student_group == "03" &
                       str_detect(district_name,"Stamford")][
  ][,.(english=mean(as.numeric(ela_performance_index),na.rm=TRUE)),
          by=.(school_name,year)][
          ][,ggplot(.SD,aes(as.integer(year),
                            english,
                            color=school_name))+
              geom_line()+ 
              ggtitle("3rd Grade English")+
              ylab("Score") + 
              xlab("Year")+ 
              scale_x_continuous(breaks = c(2014,2015,2016,2017,2018))+
              scale_y_continuous(limits=c(60,80))+
              theme_bw() +
              theme(legend.position = 'none')]

p1 <- school_perf[category=="Grade" & 
                       student_group == "04" &
                       str_detect(district_name, "Stamford")][
  ][,.(english=mean(as.numeric(ela_performance_index),na.rm=TRUE)),
          by=.(school_name,year)][
          ][,ggplot(.SD,aes(as.integer(year),
                            english,
                            color=school_name))+
              geom_line()+ 
              ggtitle("4th Grade English")+
              ylab("Score") + 
              xlab("Year")+ 
              scale_x_continuous(breaks = c(2014,2015,2016,2017,2018))+
              scale_y_continuous(limits=c(60,80))+
              theme_bw()+
              theme(legend.position = 'bottom')] 

p2 <- school_perf[category=="Grade" & 
                       student_group == "05" &
                       str_detect(district_name, "Stamford")][
  ][,.(english=mean(as.numeric(ela_performance_index),na.rm=TRUE)),
          by=.(school_name,year)][
          ][,ggplot(.SD,aes(as.integer(year),
                            english,
                            color=school_name))+
              geom_line()+ 
              ggtitle("5th Grade English")+
              ylab("Score") + 
              xlab("Year")+ 
              scale_x_continuous(breaks = c(2014,2015,2016,2017,2018))+
              scale_y_continuous(limits=c(60,80))+
              theme_bw() +
              theme(legend.position = 'none')] 

p +p1 + p2 + plot_layout(ncol=3)
```

```{r}
hn <- school_perf[str_detect(district_name, "Stamford") &
                        !str_detect(school_name,"High") &
                        !str_detect(school_name,"Middle") &
                    student_group=="High Needs"]

total <- school_perf[str_detect(district_name, "Stamford") &
                        !str_detect(school_name,"High") &
                        !str_detect(school_name,"Middle") &
                       student_group=="School"]

p <- hn[total, on = c("year","school_name")][, hn_ela:=as.numeric(ela_count)/as.numeric(i.ela_count)][,ggplot(.SD,aes(year,hn_ela,color=school_name,group=school_name))+geom_line() + ggtitle("English High Needs")+  ylab("Percentage") + 
              xlab("Year")+ 
              scale_x_discrete(breaks = c(2014,2015,2016,2017,2018))+
              theme_bw() +
              theme(legend.position = 'bottom')]

p1 <- hn[total, on = c("year","school_name")][, hn_math:=as.numeric(math_count)/as.numeric(i.math_count)][,ggplot(.SD,aes(year,hn_math,color=school_name,group=school_name))+geom_line()+ ggtitle("Math High Needs")+  ylab("Percentage") + 
              xlab("Year")+ 
              scale_x_discrete(breaks = c(2014,2015,2016,2017,2018))+
              theme_bw() +
              theme(legend.position = 'none')]

p +p1 + plot_layout(ncol=2)
```

